Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-08-30 pnas/Towards.tex
no message

Crucially, it can be shown that reliable detection of individual particles is impossible below a certain critical SNR. This fact has been recognized early on by the cryo-EM community. In particular, in an influential paper from 1995, Richard Henderson~\cite{henderson1995limitations} investigates the following questions:
\begin{quote}
	\emph{For the purposes of this review, I would like to ask the question: what is the smallest size of free-standing molecule whose structure can in principle be determined by phase-contrast electron microscopy? Given what has already been demonstrated in published work, this reduces to the question: what is the smallest size of molecule for which it is possible to determine from images of unstained molecules the five parameters needed to define accurately its orientation (three parameters) and position (two parameters) so that averaging can be performed?}
\end{quote}
In that paper and in others that followed (e.g.,~\cite{glaeser1999electron}), it was established that particle picking is impossible for molecules below a certain weight (below $\sim$50 kDa). 
Joachim Frank voices a similar observation in his 2017 Nobel prize lecture: ``\emph{Using the ribosome as an example, it became clear from the formula we obtained that the single-particle approach to structure research was indeed feasible for molecules of sufficient size: Particle Size > 3/[Contrast$^2$ $\times$ Resolution (as length) $\times$ Critical Electron Dose]}''~\cite{frank2018single}. 
\TODO{As these two leaders of the cryo-EM community point out,} it is impossible to reconstruct small molecules by any of the existing computational pipelines for single particle analysis in cryo--EM, because the particles themselves cannot be picked from the micrographs.

@@

Beyond this 2-D proof of concept, we look toward \mbox{3-D} reconstruction as well. Zvi Kam~\cite{kam1980reconstruction} first proposed autocorrelation analysis for \mbox{3-D} reconstruction, under the assumption of perfect particle picking: his method used autocorrelations of the picked particles. In contrast, we derive the mathematical relation between the autocorrelations of the micrographs as a whole and the \mbox{3-D} volume, under some simplifying conditions.
We show a few numerical examples and outline the future developments required to make this method applicable on real data. \TODO{This paragraph will be changed according to the final results we will show}

>>>>

Crucially, it can be shown that reliable detection of individual particles is impossible below a certain critical SNR. This fact has been recognized early on by the cryo-EM community. In particular, in an influential paper from 1995, Richard Henderson~\cite{henderson1995limitations} investigates the following questions:
\begin{myquote}{0.15in}
	\emph{For the purposes of this review, I would like to ask the question: what is the smallest size of free-standing molecule whose structure can in principle be determined by phase-contrast electron microscopy? Given what has already been demonstrated in published work, this reduces to the question: what is the smallest size of molecule for which it is possible to determine from images of unstained molecules the five parameters needed to define accurately its orientation (three parameters) and position (two parameters) so that averaging can be performed?}
\end{myquote}
%\begin{quote}
%	\emph{For the purposes of this review, I would like to ask the question: what is the smallest size of free-standing molecule whose structure can in principle be determined by phase-contrast electron microscopy? Given what has already been demonstrated in published work, this reduces to the question: what is the smallest size of molecule for which it is possible to determine from images of unstained molecules the five parameters needed to define accurately its orientation (three parameters) and position (two parameters) so that averaging can be performed?}
%\end{quote}
In that paper and in others that followed (e.g.,~\cite{glaeser1999electron}), it was established that particle picking is impossible for molecules below a certain weight (below $\sim$50 kDa). 
Joachim Frank voices a similar observation in his 2017 Nobel prize lecture: ``\emph{Using the ribosome as an example, it became clear from the formula we obtained that the single-particle approach to structure research was indeed feasible for molecules of sufficient size: Particle Size > 3/[Contrast$^2$ $\times$ Resolution (as length) $\times$ Critical Electron Dose]}''~\cite{frank2018single}. 
\TODO{As these two leaders of the cryo-EM community point out,} it is impossible to reconstruct small molecules by any of the existing computational pipelines for single particle analysis in cryo--EM, because the particles themselves cannot be picked from the micrographs.

@@

Beyond this 2-D proof of concept, we look toward \mbox{3-D} reconstruction as well. Zvi Kam~\cite{kam1980reconstruction} first proposed autocorrelation analysis for \mbox{3-D} reconstruction, under the assumption of perfect particle picking: his method used autocorrelations of the picked, perfectly centered, particles. In contrast, we derive the mathematical relation between the autocorrelations of the micrographs as a whole and the \mbox{3-D} volume, under some simplifying conditions.
We show a few numerical examples and outline the future developments required to make this method applicable on real data.
